Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-06 Manuscript/manuscript.tex

Although scaling of the partial charges improved 
the head group order parameter response and ion binding affinity,
it at the same time deteriorated certain membrane properties; 
namely the area per lipid decreased below the experimental value. 
After directly scaling the partial charges of the head group, the glycerol
backbone, and carbonyls with a factor of 0.8, 
the area per lipid in the simulation drops approximately to 60~\AA$^2$,
which is smaller than the experimental value of 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value from the original Lipid14 model of $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region
due a reduced head group polarity after scaling charges. 
This effect is compensated
by reducing the effective radii of atoms with the scaled charges.
This is in practice done by changing the parameters $\sigma$ in the Lennard-Jones potential 
in a similar way as was done previously for the ECC ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing the $\sigma$ parameters of affected atoms by a factor of $f_\sigma = 0.89$, 
the area per molecule is restored
back, being very close to the experimental value (Table~\ref{tab:apls}).

@@

To quantify the ion binding affinities to a lipid bilayer, we
calculate the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
Such a quantity compares the adsorption of ions to the adsorption of water molecules 
at the interface without the necessity of defining a Gibbs dividing plane 
between the membrane interior and the water bulk region. 
In our simulations we only assume that the interface is located between the ion-free hydrophobic interior
of the lipid bilayer and the aqueous region far from the membrane.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
Such a setup and the above definition of bulk ion concentration provide a simple relation for the normalized surface excess $\Gamma_i^w$ for simulations of lipid bilayers, 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) .
\end{equation}
Here, $n_w$ and $n_i$ are the total numbers of water molecules and ions in the system,
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations
consisted of 128 POPC molecules and approximately 50~SPC/E~water~\cite{Berendsen1987} molecules per each lipid
in an orthorhombic simulation box with periodic boundary conditions. 
The chosen SPC/E water model, used also in the previous parametrization of 
ECC-ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014},
has a lowered dielectric constant, consistent with the ECC concept~\cite{leontyev11, leontyev14}.
Results from simulations with other water models 
(OPC \cite{Izadi14}, OPC3~\cite{Izadi16}, 
TIP3P~\cite{jorgensen83},
TIP3p-FB and TIP4p-FB~\cite{Wang2014},
and TIP4p/2005 \cite{Abascal2005}), performed to check for robustness of our approach, are available in SI.
Sodium, calcium, and chloride 
were modeled as ECC-ions with parameters from \citenum{jungwirth17-new-paper-to-be-published, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Scaled charges and van der Waals radii for atoms forming the POPC lipid were derived here, starting from the Lipid14 force field.
For comparison, simulation with unscaled charges were perfomed using the Lipid14 model used the TIP3p water model \cite{jorgensen83} and the ion model by
Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ions and TIP3P~\cite{jorgensen83}
water models reported in \cite{catte16} 
were taken directly from \cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings employed in this work are 
summarized in Table \ref{tbl:mdpar}. 
The simulatied trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The systems were simulated with the ECC-lipids model for POPC 
and SPC/E model for water using the same setup.
In these simulations ECC was also applied to the cationic surfactant 
in the same way as to POPC, i.e.,
all partial charges were scaled with the same factor as for ECC-lipids, i.e. $f_q=0.8$, 
and the same atom types with reduced $\sigma$ parameters from ECC-lipids were used. 
For comparison, simulations with unscaled charges were also performed.

@@

In conclusion, the results suggest that calcium ions bind specifically to phosphate
oxygens, occasionally interacting also with carbonyls. This is in good agreement
with previous conclusions from several experimental and computational
studies~\cite{hauser76,hauser78,herbette84,cevc90,binder02}, but
suggesting a lower relative binding affinity to the carbonyls than inferred from previous MD simulation
studies~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.
Sodium ions also interact primarily with phosphate oxygens of the POPC, 
but in contrast to calcium, the interactions purely with carbonyls are also significant.
The physiological relevance of these details of sodium interactions is, however, unclear
due to its overal very weak binding affinity.

@@

Since the same experimental data are reproduced by the present ECC-lipids model,
we can use the model to provide a more detailed interepration of the binding stoichiometry than
the previously used simple binding models~\cite{altenbach84,macdonald87}. To directly evaluate the
stoichiometry from simulations we calculated the relative propensities of \ce{Ca^{2+}}:POPC
complexes with various stoichiometries by counting the individual lipid
molecules within the distance $0.3\,\mathrm{nm}$ from each \ce{Ca^{2+}} ion,
as already done in the previous section~\ref{sec:affinity}.
The results from POPC simulation with a 285~mM concetration of CaCl$_2$ in
Fig.~\ref{fig:cacl_complexes} demonstrate the largest propensity (41\%) for the
ternary complex with the \ce{Ca^{2+}}:POPC stoichiometry of 1:2, but 
the probabilities of complexes with the stoichimetries of 1:1~(25\%)~and
1:3~(34\%)~are only slightly lower. This
suggests a more complex binding model than the previous simple ternary complex model.
%The observed propensities of
%$41\%$ for two lipids,
%$25\%$ for one lipid, and
%$34\%$ for three lipids 
With a broad brushstroke, the simulation data can, however, be viewed such that
one calcium binds to two lipids on average,
because the probabilities of the complexes with 1 or 3 lipids are almost equal to each other 
(and complexes with more than three lipids per one calcium ion were not observed).  
This probably explains why the simple the ternary complex model fitts the
experimental data, as well as the ECC-lipids simulation results, rather well
(see Fig.~S\ref{fig:cacl-bind} in SI and its caption for details).

@@

In contrast, at least an order of magnitude faster bound/unbound calcium exchange is observed within the
ECC-lipids model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are 
shorter than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$.
The longest observed residence time is now $141\,\mathrm{ns}$, 
which is well below the total length of the simulation used for analysis, i.e., 200~ns.
The exchange of \ce{Na+} ions at the POPC membrane is
another order of magnitude faster, yielding 90\%
of the residence times smaller than~$1\,\mathrm{ns}$, with
the longest residence time being~$6\,\mathrm{ns}$. 
Note that the ECC-lipids calcium results are in line with the experimental estimate that 
the residence time of \ce{Ca^{2+}} at each individual PC headgroup 
is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}.

>>>>

Although scaling of the partial charges improved 
the head group order parameter response and ion binding affinity,
it at the same time deteriorated certain membrane properties; 
namely the area per lipid decreased below the experimental value. 
After directly scaling the partial charges of the head group, the glycerol
backbone, and carbonyls with a factor of 0.8, 
the area per lipid in the simulation drops approximately to 60~\AA$^2$,
which is smaller than the experimental value of 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value from the original Lipid14 model of $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region
due a reduced head group polarity after scaling charges. 
This effect is compensated
by reducing the effective radii of atoms with the scaled charges.
This is in practice done by changing the parameters $\sigma$ in the Lennard-Jones potential 
in a similar way as was done previously for the ECC-ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing the $\sigma$ parameters of affected atoms by a factor of $f_\sigma = 0.89$, 
the area per molecule is restored
back, being very close to the experimental value (Table~\ref{tab:apls}).

@@

To quantify the ion binding affinities to a lipid bilayer, we
calculate the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
Such a quantity compares the adsorption of ions to the adsorption of water molecules 
at the interface without the necessity of defining a Gibbs dividing plane 
between the membrane interior and the water bulk region. 
In our simulations we only assume that the interface is located between the ion-free hydrophobic interior
of the lipid bilayer and the aqueous region far from the membrane.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
Such a setup and the above definition of bulk ion concentration provide a simple relation for the relative surface excess $\Gamma_i^w$ for simulations of lipid bilayers, 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) .
\end{equation}
Here, $n_w$ and $n_i$ are the total numbers of water molecules and ions in the system,
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations
consisted of 128 POPC molecules and approximately 50~SPC/E~water~\cite{Berendsen1987} molecules per each lipid
in an orthorhombic simulation box with periodic boundary conditions. 
The chosen SPC/E water model, used also in the previous parametrization of 
ECC-ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014},
has a lowered dielectric constant, consistent with the ECC concept~\cite{leontyev11, leontyev14}.
Results from simulations with other water models 
(OPC \cite{Izadi14}, OPC3~\cite{Izadi16}, 
TIP3P~\cite{jorgensen83},
TIP3p-FB and TIP4p-FB~\cite{Wang2014},
and TIP4p/2005 \cite{Abascal2005}), performed to check for robustness of our approach, are available in SI.
Sodium, calcium, and chloride 
were modeled as ECC-ions with parameters from \citenum{jungwirth17-new-paper-to-be-published, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Scaled charges and van der Waals radii for atoms forming the POPC lipid were derived here, starting from the Lipid14 force field.
For comparison, simulations with unscaled charges were perfomed using the Lipid14 model, the TIP3p model for water \cite{jorgensen83} and the ion model by
Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ions and TIP3P~\cite{jorgensen83}
water models reported in \cite{catte16} 
were taken directly from \cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings employed in this work are 
summarized in Table \ref{tbl:mdpar}. 
The simulatied trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The systems were simulated with the ECC-lipids model for POPC 
and SPC/E model for water using the same setup.
In these simulations ECC was also applied to the cationic surfactant 
in the same way as to POPC, i.e.,
all partial charges were scaled with the same factor as for ECC-lipids, i.e. $f_q=0.8$, 
and the same atom types with reduced $\sigma$ parameters from ECC-lipids were used. 
\todo{Joe: I think that the followin is already said in the beginning of this section -- Lipid14 simulaitons.}
For comparison, simulations with unscaled charges were also performed.

@@

In conclusion, the results suggest that calcium ions bind specifically to phosphate
oxygens, occasionally interacting also with carbonyls. This is in good agreement
with previous conclusions from several experimental and computational
studies~\cite{hauser76,hauser78,herbette84,cevc90,binder02}, but
suggesting a lower relative binding affinity to the carbonyls than inferred from previous MD simulation
studies~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.
Sodium ions also interact primarily with phosphate oxygens of the POPC, 
but in contrast to calcium, the interactions purely with carbonyls are also significant.
The physiological relevance of these details of sodium interactions is, however, unclear
due to its overall very weak binding affinity.

@@

Since the same experimental data are reproduced by the present ECC-lipids model,
we can use the model to provide a more detailed interepration of the binding stoichiometry than
the previously used simple binding models~\cite{altenbach84,macdonald87}. To directly evaluate the
stoichiometry from simulations we calculated the relative propensities of \ce{Ca^{2+}}:POPC
complexes with various stoichiometries by counting the individual lipid
molecules within the distance $0.3\,\mathrm{nm}$ from each \ce{Ca^{2+}} ion,
as already done in the previous section~\ref{sec:affinity}.
The results from POPC simulation with a 285~mM concetration of CaCl$_2$ in
Fig.~\ref{fig:cacl_complexes} demonstrate the largest propensity (41\%) for the
ternary complex with the \ce{Ca^{2+}}:POPC stoichiometry of 1:2, but 
the probabilities of complexes with the stoichimetries of 1:1~(25\%)~and
1:3~(34\%)~are only slightly lower. This
suggests a more complex binding model than the previous simple ternary complex model.
%The observed propensities of
%$41\%$ for two lipids,
%$25\%$ for one lipid, and
%$34\%$ for three lipids 
With a broad brushstroke, the simulation data can, however, be viewed such that
one calcium binds to two lipids on average,
because the probabilities of the complexes with 1 or 3 lipids are almost equal to each other 
(and complexes with more than three lipids per one calcium ion were not observed).  
This probably explains why the simple the ternary complex model fits the
experimental data, as well as the ECC-lipids simulation results, rather well
(see Fig.~S\ref{fig:cacl-bind} in SI and its caption for details).

@@

In contrast, at least an order of magnitude faster bound/unbound calcium exchange is observed within the
ECC-lipids model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are 
shorter than $60\,\mathrm{ns}$, % exactly $53\,\mathrm{ns}$.
The longest observed residence time is now $141\,\mathrm{ns}$, 
which is well below the total length of the simulation used for analysis, i.e., 200~ns.
The exchange of \ce{Na+} ions at the POPC membrane is
another order of magnitude faster, yielding 90\%
of the residence times smaller than~$1\,\mathrm{ns}$, with
the longest residence time being~$6\,\mathrm{ns}$. 
\todo{Joe: the following is a bit more subtle argument. 
I deem it not necessary to include it in the paper, however, 
the experimental estimate of less tan 10us holds likely also for Charmm36/ECC-ions, as stated before.}
Note that the ECC-lipids calcium results are in line with the experimental estimate that 
the residence time of \ce{Ca^{2+}} at each individual PC headgroup 
is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}.
